An assessment of the biodiversity - ecosystem function relationship in southern African woodlands

### 2020-04-09 manuscript/regional_befr_manuscript.tex
Went through some comments

\section*{Acknowledgements}

@@

\section*{Biosketch}

@@

\section*{Abstract}

\textbf{Aim:} Positive correlations between tree biodiversity and ecosystem function have been widely documented, but the nature of the relationship \rchange{in highly disturbed and ecophysiologically stressful systems is less clear}{in southern African savanna/woodlands, which experience high levels of disturbance and ecophysiological stress, is less clear}. Here, we explore the relationship between tree biodiversity and aboveground biomass across southern African savannas and woodlands, with respect to gradients in stem density, resource availability \rnew{and disturbance through fire}, to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.

@@

\section*{Introduction}

Scientific interest in the relationship between species richness and ecosystem function springs from both an interest in the factors which structure ecological communities \citep{}, and a more applied interest in determining the effect of global biodiversity loss on ecosystem form and function \citep{}. Numerous studies have shown relationships between biodiversity and ecosystem function (e.g. \citealt{Liang2016, Hooper2012, Cardinale2009}). The strength and direction of these observed Biodiversity-Ecosystem Function Relationships (BEFRs) varies depending on the ecosystem being studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem function \citep{Liang2016}. Over the past decade, many observational studies of the BEFR have been conducted, mostly in wet tropical and temperate forests, and grasslands \citep{Chen2011}. These studies support early findings from small scale experimental studies predominantly in grassland patches, which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.

Ecosystem functions can be defined in broad terms as rate processes and properties of ecosystems which describe the degree of biotic activity within an ecosystem \citep{Jax2005}. This includes basic processes of primary production such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect measures of function such as resistance of productivity to disturbance, and further to ecosystem properties which themselves influence process, such as trophic complexity and total vegetative biomass. The frequently reported and intuitive relationship between biodiversity and ecosystem function invokes three main mechanisms which drive the relationship \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater diversity fill a greater breadth of realised niche space and avoid competition due to differences in their traits, 2) selection effects, whereby communities with greater diversity are more likely to include a species which contributes highly to the measured ecosystem function, and 3) facilitation effects, whereby communities with greater diversity are more likely to include combinations of species which together increase the others' functional contribution.

@@

A small number of studies in SAWs, all of which were restricted in the spatial scope to a small region of miombo woodland, have found that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}. The results of these fine scale studies concur with similar studies in other biomes \citep{}. Studies of the BEFR often find that at fine scales, diversity shows a strong effect on ecosystem function, but at broad scales diversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}. Due to the highly variable environmental conditions within which SAWs occur \citep{Frost1996}, and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and above ground woody biomass \citep{}, it is important to sample across geographic and environmental gradients to gain understanding of the spatial variation in the relationship between diversity and biomass.

@@

In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014, Michaletz2018} and species diversity independently \citep{Spasojevic2014}. It is important therefore to account for climatic factors and understand how they interact with biomass and diversity to effectively model and correctly attribute the effects of biodiversity on woody biomass in analyses at broad spatial scales. \citet{Sankaran2005} used data from 854 African woodland field sites to show that \rnew{below a threshold of \textapprox{}650 mm MAP}, precipitation sets sets the upper limit for woody cover in savannas, which is positively correlated with biomass \citep{Chisholm2013, Prado-Junior2016}. Similarly, \citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution and abundance evenness in a wet Panamanian tropical forest.

\citet{Solbrig1996} writes that SAWs possess structurally diverse tree canopies, with trees occupying distinct layers of the canopy at different growth stages and among species. This structural diversity may be one mechanism through which diversity influences woody biomass. \citet{Kunz2019} found that crown complementarity and crown plasticity both increased with species richness in a seasonally dry subtropical forest. They also found that trees growing in species rich neighbourhoods exhibited enhanced biomass production. Occupation of multiple canopy layers allows a more full canopy with a greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area via a form of niche complementarity. This mechanism however, which has been supported by experiments and observational studies in temperate and wet tropical ecosystems \citep{Hardiman2011, Stark2012}, may not be relevant in savannas, which are structured by disturbance rather than competition. Instead, disturbance history may override the effects of species diversity on structural diversity nullifying the effects of species diversity on structural diversity.

High levels of disturbance in SAWs may moderate the observable BEFR through its effect on ecosystem composition. Fire disturbance in forests has been linked to abundance dependent mortality among smaller tree stems \citep{Roques2001, Staver2009, Bond2005}. Some species in the regional species pool may be excluded from woodland plots with high levels of disturbance if they are unable to escape the fire bottleneck and grow to become a large tree. Selection effects may therefore be more important in maximising ecosystem function in disturbance prone woodlands. If the regional species pool contains a large number of species, it is more likely that one of them will possess the necessary growth strategy to grow to a large tree with high AGB under an intense disturbance regime.

In this study, we made the first known regional estimation of the Biodiversity-Ecosystem Function Relationship across southern African savannas and woodlands (SAWs), using inventory plots which span environmental and biogeographical gradients (\autoref{clust_map}). We used aboveground woody biomass of trees and compared the relative effects of tree species diversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation, namely water availability, energy input and soil fertility. We also investigated the potential moderating effects of environmental covariates on the relationship between tree species diversity and biomass. We incorporated vegetation type based on major tree species compositional units as a factor in our analyses to understand how species composition as well as species diversity affected ecosystem function and assess the generality of our results. We used Structural Equation Modelling (SEM) and path analysis as a preferred method to simultaneously account for environmental and biotic factors, \rchange{which may interact their effect}{which may have interacting effects} on ecosystem structure and therefore biomass. Initially, we made three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) \rchange{the effect size}{the strenghth of the effect} of species diversity on woody biomass will increase with stem density, due to an increased importance of niche complementarity as competition increases, and (3) tree species diversity will increase tree structural diversity, which will provide an indirect path by which tree diversity increases woody biomass.

\section*{Materials and methods}

\subsection*{Study location}

@@

\subsection*{Data collection}

@@

Climatic data were collected from the ECMWF ERA5 dataset, generated using Copernicus Climate Change Service Information \citep{ERA5}. Values of \rchange{Mean Annual Temperature (MAT)}{MAT} and \rchange{Mean Annual Precipitation (MAP)}{MAP} were calculated from daily data between 2000 and 2018, then averaged across years to provide a single mean annual estimate per plot. Temperature seasonality (TS) and precipitation seasonality (PS) were both calculated as the mean of the coefficient of variation of daily MAT and MAP, respectively, for each of the 18 years of available data. Soil fertility data was extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot centre \citep{Hengl2017}. We extracted Cation Exchange Capacity (CEC), percentage soil organic carbon by volume (Org. C \%), and percentage soil sand content by volume (Sand \%). These data are a modelled product derived from various remotely sensed and directly measured data sources.

@@

\subsection*{Data analysis}
Estimated tree species richness was calculated for each plot using \verb|ChaoRichness()| from the \verb|iNEXT| package in R \citep{Hsieh2016}. This procedure extrapolates a species rarefaction curve to its predicted asymptote and uses this value as its estimated species richness value. Extrapolated species richness accounts for variation in plot size (0.1-10 ha) and therefore sampling effort among plots. Larger plots will tend to encompass more individuals, and therefore more species \citep{Dengler2009}.

To measure tree species abundance evenness, the Shannon Equitability index ($E_{H'}$) \citep{Smith1996} (\autoref{shannon_equit}) was calculated as the ratio of estimated Shannon diversity to the natural log of estimated species richness. Abundance evenness allows for greater niche complementarity at small scales due to an increased spatial heterogeneity of functional traits. We calculated tree structural diversity for each plot by calculating the coefficient of variation of DBH (DBH CV) and tree height (Height CV).

@@

\subsubsection*{Vegetation clusters}

@@

\subsubsection*{Structural Equation Modelling}

Structural Equation Models (SEM) investigated the determinants of AGB. All SEMs were constructed and analysed in the \verb|lavaan| package \citep{lavaan} in R version 3.6.0 \citep{R2019}. SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lee2007}. A key aspect to our decision to use SEMs is that they can explicitly model and partition variance to indirect effects, which is challenging in standard multiple regressions. Using SEMs also allowed us to describe theoretical latent constructs which have been suggested to act upon diversity and biomass/productivity in previous studies despite these factors not having single observable measures in our dataset. Structural equation modelling is also necessary to properly account for potential feedback mechanisms between aspects of climate and tree species diversity, which could otherwise increase the chances of Type I error and wrongly attribute inference due to covariance of explanatory variables when using conventional regression analyses \citep{Nachtigall2003}.

@@

We assessed the role of structural diversity and species diversity in determining AGB via a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity. To account for variation in stem density which may covary with species diversity we also included it as an predictor in our model. To explore variation in the model among woodland vegetation types, we fit the model both at the regional scale and for each vegetation cluster separately. We compared unstandardised path coefficients among these vegetation cluster scale models to understand the effect that vegetation type has on the relationship between tree species diversity, structural diversity, stem density and AGB. Path coefficients show the effect of a path with other paths of inference held constant. Models were estimated using the ``MLM'' estimator, because it is robust to multivariate non-normality \citep{Shapiro1983}. Model fit was evaluated using the robust Comparative Fit Index (CFI), the robust Tucker Lewis Index (TLI), the Root Mean Squared Error (RMSEA) and the R\textsuperscript{2} coefficient of determination for AGB. We critically assess model fit in each case, taking into consideration the recommendations of \citet{Hu1999} which define threshold values of acceptability for these model fit indices: CFI >0.85, TLI >0.85, RMSEA <0.15, alongside our judgement of the model estimates.

@@

\section*{Results}

@@

\subsection*{Structural and species diversity models}

@@

\subsection*{Variation among vegetation types}

@@

\subsection*{Moderation of Diversity-AGB relationship by stem density}

@@

\subsection*{Environmental covariates and diversity}

A model incorporating the latent variables of moisture availability and soil fertility showed that the total effect of species diversity on biomass was greater than that of both moisture availability and soil fertility (\autoref{full_mod}). Surprisingly, the direct effects of moisture availability and soil fertility on biomass were negligible, with nearly all of their observed effect on AGB coming from the indirect path via species diversity (moisture: \rgmbd{}, soil: \rgsbd{}). MAP and temperature seasonality (TS) had the greatest contributions to the latent variable of moisture availability. Moisture availability and soil fertility also had negligible direct effects on stem density.  Model fit was acceptable: CFI = \fmcfi{}, TLI = \fmtli{}, and RMSEA = \fmrmsea{}, R\textsuperscript{2} of AGB = \fmrsq{}.

@@

% \subsection*{Moderation effects of moisture availability and soil}
% 
% To address the hypothesis (H\textsubscript{1}) that more arid plots and plots with less fertile soil will show a stronger positive effect of tree species richness on above ground woody biomass, we fit separate linear multiple regressions with the latent variables of moisture availability and soil fertility as moderators on the relationship between species diversity and AGB.
% 
% Both soil fertility and moisture availability had small positive interaction effects on the strength of the relationship between tree species diversity and biomass (\autoref{int_plots}). The regression coefficient for the interaction effect of tree species diversity and moisture availability was significant (\moderp{}), while the interaction term of tree species diversity and soil fertility was marginally significant (\moders{}). As moisture availability increased, the effect of tree species diversity on AGB became stronger. See \autoref{mois_div_int_mod} and \autoref{soil_div_int_mod} for full regression tables.
% 
% 
% \begin{figure}[H]
% \centering
% 	\subfloat[]{
% 	\includegraphics[width=0.48\textwidth]{mois_int}
% 	}
% 	\subfloat[]{
% 	\includegraphics[width=0.48\textwidth]{soil_int}
% 	}
% 	\caption{Scatter plots showing variation in the relationship between tree species diversity and AGB with moisture availability (a) and soil fertility (b). All variables are standardised and centred to a mean of zero and a standard deviation of 1. Lines of best fit are drawn for standard deviations of moisture availability and soil fertility between -2 and +2. Moisture availability, tree species diversity and soil fertility are latent variables comprised of the same observed variable loadings.}
% 	\label{int_plots}
% \end{figure}

\section*{Discussion}

@@

In this study, we assessed the importance of [a] tree species richness, [b] tree structural diversity, [c] resource availability, i.e. moisture availability and soil fertility, [d] stem density and their interactions on above ground woody biomass (AGB) across SAWs, using a network of \nplots{} woodland survey plots. Using latent variables and Structural Equation Modelling (SEM), we found support for a general positive relationship between tree species diversity and AGB, with an indirect influence of tree species diversity on AGB via structural diversity (H\textsubscript{1}). We found that the effect size of tree species diversity on AGB increased with stem density (H\textsubscript{2}).  Tree diversity, structural diversity and stem density accounted for \srsq{}\% of the variation in AGB across the region, while models for specific vegetation types showed even greater explanatory power in some cases (\autoref{struc_model_fit_clust_stats}). The strongest effect on AGB was that of stem density. Interestingly, when the effects of tree species diversity, structural diversity and stem density were controlled for, we found little evidence of a direct effect of resource availability, in the form of moisture or soil fertility, on AGB (H\textsubscript{3}).

\subsection*{Inter-related effects of tree species and structural diversity on AGB}

We found a consistent positive effect of tree species diversity on AGB across all models in this study. Within SAWs we therefore find support that higher tree species richness and evenness causes higher woody AGB. This finding is in agreement with many other studies across different ecosystems and biomes, supporting that there is a generalisable positive association between diversity and ecosystem function \citep{Liang2016, Cardinale2009}. Our study provides a novel dissection of the mechanisms underlying this relationship, particularly in the context of SAWs, a disturbance-structured and poorly studied ecological system.

@@

\subsection*{Effects of moisture availability and soil fertility}

@@

\subsection*{Vegetation type specific responses}

@@

Core miombo and marginal miombo woodland vegetation exhibited a small negative direct effect of tree species diversity on AGB, while the total effect, incorporating the indirect effect via  structural diversity, remained positive in these vegetation types. Compared to Baikiaea and Mopane woodlands, miombo woodlands have higher median tree species richness. Baikiaea and Mopane woodlands are also dominated by fewer tree species, notably \textit{Baikiaea plurijuga} in Baikiaea woodlands and \textit{Colophospermum mopane} in Mopane woodlands which often produce large canopy dominating trees. We postulate that this negative effect of tree species richness on AGB in miombo woodlands may be due to an increase in interspecific competition through canopy crowding, but that this effect is not present in Baikiaea and Mopane woodlands, where the woodland canopy is dominated often by a single species. Higher functional redundancy among tree species in miombo woodlands may lead to smaller trees with lower AGB in the most diverse plots, more resembling thicket vegetation. Again, these highly diverse plots in miombo woodlands may be the result of disturbance which can promote a mosaic of woodland of different successional stages and stem densities. Alternatively, this small negative direct effect may be an artefact of particularly noisy data, especially given that the overall effect of diversity on AGB is positive.

@@

\subsection*{Conclusion}

@@

\section*{Data accessibility statement}

\section*{Tables}

\section*{Figure legends and embedded figures}

\section*{Appendix 1 - Data cleaning process} \label{appendixa}

@@

\section*{Appendix 2 - Estimation of DBH via tree taper} \label{appendixb}

@@

\section*{Appendix 3 - Frequency distribution of observed variables} \label{appendixc}

@@

\section*{Appendix 4 - Table of correlation fit statistics} \label{appendixd}

>>>>

\section{Acknowledgements}

@@

\section{Biosketch}

@@

\section{Abstract}

\textbf{Aim:} Positive correlations between tree biodiversity and ecosystem function have been widely documented, but the nature of the relationship \rchange{in highly disturbed and ecophysiologically stressful systems is less clear}{in southern African savanna/woodlands, which experience high levels of disturbance and ecophysiological stress, is less clear}. Here, we explore the relationship between tree species diversity and aboveground biomass across southern African savannas and woodlands, with respect to gradients in stem density, resource availability \rnew{and disturbance through fire}, to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.

@@

\section{Introduction}

Scientific interest in the relationship between species diversity and ecosystem function springs from both an interest in the factors which structure ecological communities \citep{}, and a more applied interest in determining the effect of global biodiversity loss on ecosystem form and function \citep{}. Numerous studies have shown relationships between biodiversity and ecosystem function (e.g. \citealt{Liang2016, Hooper2012, Cardinale2009}). The strength and direction of these observed Biodiversity-Ecosystem Function Relationships (BEFRs) varies depending on the ecosystem being studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem function \citep{Liang2016}. Over the past decade, many observational studies of the BEFR have been conducted, mostly in wet tropical and temperate forests, and grasslands \citep{Chen2011}. These studies support early findings from small scale experimental studies predominantly in grassland patches, which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.

Ecosystem functions can be defined in broad terms as rate processes and properties of ecosystems which describe the degree of biotic activity within an ecosystem \citep{Jax2005}. This includes basic processes of primary production such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect measures of function such as resistance of productivity to disturbance, and further to ecosystem properties which themselves influence process, such as trophic complexity and total vegetative biomass. The frequently reported and intuitive relationship between biodiversity and ecosystem function invokes three main mechanisms which drive the relationship \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater biodiversity fill a greater breadth of realised niche space and avoid competition due to differences in their traits, 2) selection effects, whereby communities with greater biodiversity are more likely to include a species which contributes highly to the measured ecosystem function, and 3) facilitation effects, whereby communities with greater biodiversity are more likely to include combinations of species which together increase the others' functional contribution.

@@

A small number of studies in SAWs, all of which were restricted in the spatial scope to a small region of miombo woodland, have found that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}. The results of these fine scale studies concur with similar studies in other biomes \citep{}. Studies of the BEFR often find that at fine scales, biodiversity shows a strong effect on ecosystem function, but at broad scales biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}. Due to the highly variable environmental conditions within which SAWs occur \citep{Frost1996}, \rnew{with wide variation in precipitation, diurnal and annual temperature range} and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and above ground woody biomass \citep{}, it is important to sample across geographic and environmental gradients to gain understanding of the spatial variation in the relationship between biodiversity and biomass.

@@

In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014, Michaletz2018} and tree species diversity independently \citep{Spasojevic2014}. It is important therefore to account for climatic factors and understand how they interact with biomass and biodiversity to effectively model and correctly attribute the effects of biodiversity on woody biomass in analyses at broad spatial scales. \citet{Sankaran2005} used data from 854 African woodland field sites to show that \rnew{below a threshold of \textapprox{}650 mm MAP}, precipitation sets sets the upper limit for woody cover in savannas, which is positively correlated with biomass \citep{Chisholm2013, Prado-Junior2016}. Similarly, \citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution and abundance evenness in a wet Panamanian tropical forest.

\citet{Solbrig1996} writes that SAWs possess structurally diverse tree canopies, with trees occupying distinct layers of the canopy at different growth stages and among species. This structural diversity may be one mechanism through which tree species diversity influences woody biomass. \citet{Kunz2019} found that crown complementarity and crown plasticity both increased with species richness in a seasonally dry subtropical forest. They also found that trees growing in species rich neighbourhoods exhibited enhanced biomass production. Occupation of multiple canopy layers allows a more full canopy with a greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area via a form of niche complementarity. This mechanism however, which has been supported by experiments and observational studies in temperate and wet tropical ecosystems \citep{Hardiman2011, Stark2012}, may not be relevant in savannas, which are structured by disturbance rather than competition. Instead, disturbance history may override the effects of tree species diversity on structural diversity nullifying the effects of tree species diversity on structural diversity.

High levels of disturbance in SAWs may moderate the observable BEFR through its effect on ecosystem composition. Fire disturbance in forests has been linked to abundance dependent mortality among smaller tree stems \citep{Roques2001, Staver2009, Bond2005}. Some species in the regional species pool may be excluded from woodland plots with high levels of disturbance if they are unable to escape the fire bottleneck and grow to become a large tree. Selection effects may therefore be more important in maximising ecosystem function in disturbance prone woodlands. If the regional species pool contains a large number of species, it is more likely that one of them will possess the necessary growth strategy to grow to a large tree with high biomass under an intense disturbance regime.

In this study, we made the first known regional estimation of the biodiversity-ecosystem function relationship across southern African savannas and woodlands (SAWs), using inventory plots which span environmental and biogeographical gradients (\autoref{clust_map}). We used aboveground woody biomass of trees and compared the relative effects of tree species diversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation, namely water availability, energy input and soil fertility. We also investigated the potential moderating effects of environmental covariates on the relationship between tree species diversity and biomass. We incorporated vegetation type via clustering of plot level tree species composition, as a factor in our analyses to understand how tree species composition as well as diversity affected ecosystem function and assess the generality of our results. We used Structural Equation Modelling (SEM) and path analysis as a preferred method to simultaneously account for environmental and biotic factors, \rchange{which may interact their effect}{which may have interacting effects} on ecosystem structure and therefore biomass. Initially, we made three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) \rchange{the effect size}{the strenghth of the effect} of tree species diversity on woody biomass will increase with plot level stem stocking density, due to an increased importance of niche complementarity as competition increases, and (3) tree species diversity will increase tree structural diversity, which will provide an indirect path by which tree diversity increases woody biomass.

\section{Materials and methods}

\subsection{Study location}

@@

\subsection{Data collection}

@@

Climatic data were collected from the ECMWF ERA5 dataset, generated using Copernicus Climate Change Service Information \citep{ERA5}. Values of \rchange{Mean Annual Temperature (MAT)}{MAT} and \rchange{Mean Annual Precipitation (MAP)}{MAP} were calculated from daily data between 2000 and 2018, then averaged across years to provide a single mean annual estimate per plot. Temperature seasonality (TS) and precipitation seasonality (PS) were both calculated as the mean of the coefficient of variation of daily mean temperature and precipitation, respectively, for each of the 18 years of available data. Soil fertility data was extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot centre \citep{Hengl2017}. We extracted Cation Exchange Capacity (CEC), percentage soil organic carbon by volume (Org. C \%), and percentage soil sand content by volume (Sand \%). These data are a modelled product derived from various remotely sensed and directly measured data sources.

@@

\subsection{Data analysis}

\subsubsection{Species diversity and structural diversity metrics}

Estimated tree species richness was calculated for each plot using \verb|ChaoRichness()| from the \verb|iNEXT| package in R \citep{Hsieh2016}. This procedure extrapolates a species rarefaction curve to its predicted asymptote and uses this value as its estimated species richness value. Extrapolated species richness accounts for variation in plot size (0.1-10 ha) and therefore sampling effort among plots. Larger plots will tend to encompass more individuals, and therefore more species \citep{Dengler2009}. To measure tree species abundance evenness, the Shannon Equitability index ($E_{H'}$) \citep{Smith1996} (\autoref{shannon_equit}) was calculated as the ratio of the estimated Shannon diversity index to the natural log of estimated species richness. Abundance evenness allows for greater niche complementarity at small scales due to an increased spatial heterogeneity of functional traits. We calculated tree structural diversity for each plot by calculating the coefficient of variation of DBH (DBH CV) and tree height (Height CV).

@@

\subsubsection{Vegetation clusters}

@@

\subsubsection{Structural Equation Modelling}

Structural Equation Models (SEM) investigated the determinants of AGB. All SEMs were constructed and analysed in the \verb|lavaan| package \citep{lavaan} in R version 3.6.0 \citep{R2019}. SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lee2007}. A key aspect to our decision to use SEMs is that they can explicitly model and partition variance to indirect effects, which is challenging in standard multiple regressions. Using SEMs also allowed us to describe theoretical latent constructs which have been suggested to act upon biodiversity and biomass/productivity in previous studies despite these factors not having single observable measures in our dataset. Structural equation modelling is also necessary to properly account for potential feedback mechanisms between aspects of climate and tree species diversity, which could otherwise increase the chances of Type I error and wrongly attribute inference due to covariance of explanatory variables when using conventional regression analyses \citep{Nachtigall2003}.

@@

We assessed the role of tree structural diversity and species diversity in determining AGB via a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity. To account for variation in stem density which may covary with species diversity we also included it as an predictor in our model. To explore variation in the model among woodland vegetation types, we fit the model both at the regional scale and for each vegetation cluster separately. We compared unstandardised path coefficients among these vegetation cluster scale models to understand the effect that vegetation type has on the relationship between tree species diversity, structural diversity, stem density and AGB. Path coefficients show the effect of a path with other paths of inference held constant. Models were estimated using the ``MLM'' estimator, because it is robust to multivariate non-normality \citep{Shapiro1983}. Model fit was evaluated using the robust Comparative Fit Index (CFI), the robust Tucker Lewis Index (TLI), the Root Mean Squared Error (RMSEA) and the R\textsuperscript{2} coefficient of determination for AGB. We critically assess model fit in each case, taking into consideration the recommendations of \citet{Hu1999} which define threshold values of acceptability for these model fit indices: CFI >0.85, TLI >0.85, RMSEA <0.15, alongside our judgement of the model estimates.

@@

\section{Results}

@@

\subsection{Structural and species diversity models}

@@

\subsection{Variation among vegetation types}

@@

\subsection{Moderation of Diversity-AGB relationship by stem density}

@@

\subsection{Environmental covariates and tree diversity}

A model incorporating the latent variables of moisture availability and soil fertility showed that the total effect of tree species diversity on biomass was greater than that of both moisture availability and soil fertility (\autoref{full_mod}). Surprisingly, the direct effects of moisture availability and soil fertility on biomass were negligible, with nearly all of their observed effect on AGB coming from the indirect path via species diversity (moisture: \rgmbd{}, soil: \rgsbd{}). MAP and temperature seasonality (TS) had the greatest contributions to the latent variable of moisture availability. Moisture availability and soil fertility also had negligible direct effects on stem density.  Model fit was acceptable: CFI = \fmcfi{}, TLI = \fmtli{}, and RMSEA = \fmrmsea{}, R\textsuperscript{2} of AGB = \fmrsq{}.

@@

% \subsection{Moderation effects of moisture availability and soil}
% 
% To address the hypothesis (H\textsubscript{1}) that more arid plots and plots with less fertile soil will show a stronger positive effect of tree species richness on above ground woody biomass, we fit separate linear multiple regressions with the latent variables of moisture availability and soil fertility as moderators on the relationship between species diversity and AGB.
% 
% Both soil fertility and moisture availability had small positive interaction effects on the strength of the relationship between tree species diversity and biomass (\autoref{int_plots}). The regression coefficient for the interaction effect of tree species diversity and moisture availability was significant (\moderp{}), while the interaction term of tree species diversity and soil fertility was marginally significant (\moders{}). As moisture availability increased, the effect of tree species diversity on AGB became stronger. See \autoref{mois_div_int_mod} and \autoref{soil_div_int_mod} for full regression tables.
% 
% 
% \begin{figure}[H]
% \centering
% 	\subfloat[]{
% 	\includegraphics[width=0.48\textwidth]{mois_int}
% 	}
% 	\subfloat[]{
% 	\includegraphics[width=0.48\textwidth]{soil_int}
% 	}
% 	\caption{Scatter plots showing variation in the relationship between tree species diversity and AGB with moisture availability (a) and soil fertility (b). All variables are standardised and centred to a mean of zero and a standard deviation of 1. Lines of best fit are drawn for standard deviations of moisture availability and soil fertility between -2 and +2. Moisture availability, tree species diversity and soil fertility are latent variables comprised of the same observed variable loadings.}
% 	\label{int_plots}
% \end{figure}

\section{Discussion}

@@

In this study, we assessed the importance of [a] tree species diversity, [b] tree structural diversity, [c] resource availability, i.e. moisture availability and soil fertility, [d] stem density and their interactions on above ground woody biomass (AGB) across SAWs, using a network of \nplots{} woodland survey plots. Using latent variables and Structural Equation Modelling (SEM), we found support for a general positive relationship between tree species diversity and AGB, with an indirect influence of tree species diversity on AGB via structural diversity (H\textsubscript{1}). We found that the effect size of tree species diversity on AGB increased with stem density (H\textsubscript{2}).  Tree diversity, structural diversity and stem density accounted for \srsq{}\% of the variation in AGB across the region, while models for specific vegetation types showed even greater explanatory power in some cases (\autoref{struc_model_fit_clust_stats}). The strongest effect on AGB was that of stem density. Interestingly, when the effects of tree species diversity, structural diversity and stem density were controlled for, we found little evidence of a direct effect of resource availability, in the form of moisture or soil fertility, on AGB (H\textsubscript{3}).

\subsection{Inter-related effects of tree species and structural diversity on AGB}

We found a consistent positive effect of tree species diversity on AGB across all models in this study. Within SAWs we therefore find support that higher tree species richness and evenness causes higher woody AGB. This finding is in agreement with many other studies across different ecosystems and biomes, supporting that there is a generalisable positive association between biodiversity and ecosystem function \citep{Liang2016, Cardinale2009}. Our study provides a novel dissection of the mechanisms underlying this relationship, particularly in the context of SAWs, a disturbance-structured and poorly studied ecological system.

@@

\subsection{Effects of moisture availability and soil fertility}

@@

\subsection{Vegetation type specific responses}

@@

Core miombo and marginal miombo woodland vegetation exhibited a small negative direct effect of tree species diversity on AGB, while the total effect, incorporating the indirect effect via  structural diversity, remained positive in these vegetation types. Compared to Baikiaea and Mopane woodlands, miombo woodlands have higher median tree species richness. Baikiaea and Mopane woodlands are also dominated by fewer tree species, notably \textit{Baikiaea plurijuga} in Baikiaea woodlands and \textit{Colophospermum mopane} in Mopane woodlands which often produce large canopy dominating trees. We postulate that this negative effect of tree species richness on AGB in miombo woodlands may be due to an increase in interspecific competition through canopy crowding, but that this effect is not present in Baikiaea and Mopane woodlands, where the woodland canopy is dominated often by a single species. Higher functional redundancy among tree species in miombo woodlands may lead to smaller trees with lower AGB in the most diverse plots, more resembling thicket vegetation. Again, these highly diverse plots in miombo woodlands may be the result of disturbance which can promote a mosaic of woodland of different successional stages and stem densities. Alternatively, this small negative direct effect may be an artefact of particularly noisy data, especially given that the overall effect of tree diversity on AGB is positive.

@@

\subsection{Conclusion}

@@

\section{Data accessibility statement}

\section{Tables}

\section{Figure legends and embedded figures}

\section{Appendix 1 - Data cleaning process} \label{appendixa}

@@

\section{Appendix 2 - Estimation of DBH via tree taper} \label{appendixb}

@@

\section{Appendix 3 - Frequency distribution of observed variables} \label{appendixc}

@@

\section{Appendix 4 - Table of correlation fit statistics} \label{appendixd}
